---
title: "01 Titer Determination (ddPCR)"
author: "Joyce Suppacher"
date: "`2026-06-14"

---

# Titer Determination

This script visualizes vector titers determined by droplet digital PCR (ddPCR).
Titer values were pre-calculated from QuantaSoft output using:
(CopiesPer20µLWell / 5) × 1000 × dilution factor = vg/mL

Run `00_setup.Rmd` first.

```{r packages, message=FALSE, warning=FALSE}
library(readODS)   # read_ods(): load ddPCR data from ODS file
library(dplyr)     # mutate(), factor()
library(ggpubr)    # ggboxplot(), stat_compare_means()
library(ggplot2)   # scale_fill_manual(), ggsave()
```

Output:
- `Titer.svg`: Boxplot of vector titers per sample with statistical comparisons

---

## 1. Load ddPCR Data

```{r load_titer}
titer_df <- read_ods(ddpcr_file, sheet = 3)

# Ensure correct column name and sample order
colnames(titer_df)[2] <- "vg per mL"
titer_df <- titer_df %>%
  mutate(Sample = factor(Sample, levels = sample_levels))
```

---

## 2. Titer Boxplot

Statistical comparison: Wilcoxon rank-sum test, Lambda as reference group.
Non-significant comparisons are not shown (hide.ns = TRUE).

```{r titer_plot}
ggboxplot(
  data       = titer_df,
  x          = "Sample",
  y          = "vg per mL",
  fill       = "Sample",
  color      = "black",
  add        = "jitter",
  add.params = list(color = "black", size = 0.5, alpha = 1)
) +
  scale_fill_manual(values = sample_colors) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.1))) +
  stat_compare_means(
    comparisons = list(
      c("lambda", "wt-fragments"),
      c("lambda", "M1"),
      c("lambda", "M2"),
      c("lambda", "M3"),
      c("lambda", "M4")
    ),
    method   = "wilcox.test",
    label    = "p.signif",
    hide.ns  = TRUE
  ) +
  theme_publication(base_size = 10) +
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(file.path(output_dir, "Titer.svg"),
       width = 90, height = 80, units = "mm")
```

---

