---
title: "02 Quality Control of Nanopore Sequencing Data"
author: "Joyce Suppacher"
date: "2026-06-14"
---

# Quality Control

This script covers:
1. Conversion of raw Nanopore BAM output to FASTQ
2. Quality assessment of raw reads (Phred score, read length, read count)
3. Read classification (mapped/unmapped, primary/multimapped)

Run `00_setup.Rmd` first.

```{r packages, message=FALSE, warning=FALSE}
library(Biostrings)       # readDNAStringSet(), width()
library(ShortRead)        # readFastq(), quality()
library(dplyr)            # mutate(), filter(), bind_rows()
library(ggpubr)           # gghistogram(), ggviolin()
library(ggplot2)          # geom_hline(), annotate(), ggsave()
library(Rsamtools)        # countBam()
library(gt)               # gt(), tab_header(), cols_label()
```

Outputs:
- `PhredScore.svg`: Mean Phred score per read per sample (violin plot)
- `ReadLength_Distr.svg`: Read length distribution per sample (histogram)
- Read count table (printed to console)

---

## 1. BAM to FASTQ Conversion

Rebasecalled Nanopore BAM output files are converted to FASTQ format using samtools.

```{bash bam_to_fastq, eval=FALSE}
# Input:   BAM files from Nanopore rebasecalling
# Output: FASTQ files per sample

SAMPLES=("lambda" "wt-fragments" "M1" "M2" "M3" "M4")
RAW_BAM_DIR="/path"
FASTQ_DIR="/path"

mkdir -p "${FASTQ_DIR}"

for SAMPLE in "${SAMPLES[@]}"; do
    echo "Converting ${SAMPLE}..."
    samtools fastq "${RAW_BAM_DIR}/${SAMPLE}.bam" > "${FASTQ_DIR}/${SAMPLE}.fastq"
done
```

---

## 2. Load FASTQ Files

```{r load_fastq}
fastq_files <- list.files(
  fastq_dir,
  pattern    = "\\.fastq$|\\.fastq\\.gz$",
  full.names = TRUE,
  recursive  = TRUE
)

# Extract sample names
sample_names_fq <- basename(fastq_files) %>%
  gsub("\\.fastq\\.gz$|\\.fastq$", "", .)
```

---

## 3. Read Count per Sample

```{r read_count}
read_count_fq <- lapply(seq_along(fastq_files), function(i) {
  reads <- readDNAStringSet(fastq_files[i], format = "fastq")
  data.frame(
    Sample      = sample_names_fq[i],
    Total_Reads = length(reads)
  )
}) %>% bind_rows() %>%
  mutate(Sample = factor(Sample, levels = sample_levels))

print(read_count_fq)
```

---

## 4. Read Length Distribution

```{r read_length}
read_length_fq <- lapply(seq_along(fastq_files), function(i) {
  reads <- readDNAStringSet(fastq_files[i], format = "fastq")
  data.frame(
    Sample      = sample_names_fq[i],
    read_length = width(reads)
  )
}) %>% bind_rows() %>%
  mutate(Sample = factor(Sample, levels = sample_levels))

gghistogram(
  data       = read_length_fq,
  x          = "read_length",
  fill       = "Sample",
  color      = "black",
  bins       = 200,
  add        = "mean",
  add.params = list(color = "black", linetype = "dashed", linewidth = 0.25),
  rug        = FALSE
) +
  scale_fill_manual(values = sample_colors) +
  coord_cartesian(xlim = c(0, 8000)) +
  facet_wrap(~ Sample, ncol = 3) +
  labs(
    title = "Read Length Distribution (FASTQ)",
    x     = "Read Length (bp)",
    y     = "Count"
  ) +
  theme_pubr() +
  theme_publication(base_size = 10) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(file.path(output_dir, "ReadLength_Distr.svg"),
       width = 180, height = 120, units = "mm")
```

---

## 5. Phred Score Distribution

Mean Phred quality score per read, averaged on linear error probability scale.

```{r phred_score}
qscore_df <- lapply(seq_along(fastq_files), function(i) {
  reads     <- readFastq(fastq_files[i])
  qual_list <- as.character(quality(quality(reads)))

  mean_qual <- vapply(qual_list, function(q) {
    if (is.null(q) || length(q) == 0) return(NA_real_)
    scores    <- utf8ToInt(q) - 33L
    probs     <- 10^(-scores / 10)
    mean_prob <- mean(probs)
    -10 * log10(mean_prob)
  }, numeric(1))

  data.frame(Sample = sample_names_fq[i], mean_qual = mean_qual)
}) %>% bind_rows() %>%
  mutate(Sample = factor(Sample, levels = sample_levels))

ggviolin(
  data       = qscore_df,
  x          = "Sample",
  y          = "mean_qual",
  fill       = "Sample",
  alpha      = 0.8,
  add        = "boxplot",
  add.params = list(width = 0.1, fill = "white")
) +
  scale_fill_manual(values = sample_colors) +
  geom_hline(yintercept = 20, linetype = "dashed", color = "red", linewidth = 0.6) +
  labs(
    title = "Phred Score per Sample",
    y     = "Phred Score per Read (Q)",
    x     = "Sample"
  ) +
  theme_pubr() +
  theme_publication(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right")

ggsave(file.path(output_dir, "PhredScore.svg"),
       width = 180, height = 120, units = "mm")
```

---

## 6. Read Classification

This section documents the classification of reads into mapped/unmapped
and primary/multimapped categories.

Note on reported units:
- Unmapped, Mapped and Primary Aligned are reported as unique reads
- Multimapped is reported as alignments (secondary + supplementary combined),
  as one read can produce multiple supplementary alignments

### 6.1 Create Classification BAM Files


```{bash create_classification_bam, eval=FALSE}
SAMPLES=("lambda" "wt-fragments" "M1" "M2" "M3" "M4")
ALN_DIR="/path"
OUT_DIR="/path"

mkdir -p "${OUT_DIR}"

for SAMPLE in "${SAMPLES[@]}"; do
    BAM="${ALN_DIR}/sorted_alignment_${SAMPLE}.bam"

    # Unmapped reads (flag 4 set)
    samtools view -f 4 -b -h "${BAM}" > "${OUT_DIR}/unmapped_${SAMPLE}.bam"

    # Mapped reads (flag 4 not set)
    samtools view -F 4 -b -h "${BAM}" > "${OUT_DIR}/mapped_${SAMPLE}.bam"

    # Primary alignments (unique mappers)
    samtools view -F 2308 -b -h "${BAM}" > "${OUT_DIR}/primary_${SAMPLE}.bam"

    # Secondary alignments (flag 256 set)
    samtools view -f 256 -b -h "${BAM}" > "${OUT_DIR}/secondary_${SAMPLE}.bam"

    # Supplementary alignments (flag 2048 set)
    samtools view -f 2048 -b -h "${BAM}" > "${OUT_DIR}/supplementary_${SAMPLE}.bam"

    # Index all
    for TYPE in unmapped mapped primary secondary supplementary; do
        samtools index "${OUT_DIR}/${TYPE}_${SAMPLE}.bam"
    done
done
```

### 6.2 Extract Read and Alignment Counts

Unmapped, Mapped and Primary Aligned are counted as unique Reads via
Read name deduplication (`sort -u`). Secondary and Supplementary are
counted as Alignments via `samtools view -c`.


```{bash count_classification, eval=FALSE}
OUT_DIR="/path"

echo "Sample,Unmapped_Reads,Mapped_Reads,Primary_Reads,Secondary_Aln,Supplementary_Aln" \
    > "${OUT_DIR}/read_classification.csv"

for SAMPLE in lambda wt-fragments M1 M2 M3 M4; do

    # Unmapped Reads (unique Read names)
    UNMAPPED=$(samtools view "${OUT_DIR}/unmapped_${SAMPLE}.bam" \
        | cut -f1 | sort -u | wc -l)

    # Mapped Reads (unique Read names)
    MAPPED=$(samtools view "${OUT_DIR}/mapped_${SAMPLE}.bam" \
        | cut -f1 | sort -u | wc -l)

    # Primary Aligned Reads (alignment count = read count for primary)
    PRIMARY=$(samtools view -c "${OUT_DIR}/primary_${SAMPLE}.bam")

    # Secondary Alignments (alignment count)
    SECONDARY=$(samtools view -c "${OUT_DIR}/secondary_${SAMPLE}.bam")

    # Supplementary Alignments (alignment count)
    SUPPLEMENTARY=$(samtools view -c "${OUT_DIR}/supplementary_${SAMPLE}.bam")

    echo "${SAMPLE},${UNMAPPED},${MAPPED},${PRIMARY},${SECONDARY},${SUPPLEMENTARY}" \
        >> "${OUT_DIR}/read_classification.csv"

done
```

### 6.3 Build Classification Table

Values were extracted via bash and compiled into a table via Excel.
Multimapped = Secondary + Supplementary Alignments.

---

