---
title: "03 Truncation Analysis"
author: "Joyce Suppacher"
date: "2026-06-14"
---

# Truncation Analysis

This script covers:
1. Alignment of FASTQ files to transgene references
2. Sorting, indexing and filtering of BAM files
3. Truncation analysis (alignment length, match count, genome integrity, coverage)

Run `00_setup.Rmd` first.

```{r packages, message=FALSE, warning=FALSE}
library(GenomicAlignments) # readGAlignments(), cigar(), start(), end(), width()
library(dplyr)             # mutate(), filter(), bind_rows(), group_by(), summarise()
library(tidyr)             # pivot_longer()
library(ggpubr)            # gghistogram(), ggscatter(), ggbarplot()
library(ggplot2)           # geom_histogram(), geom_hline(), ggsave()
library(purrr)             # map_dfr()
```

Outputs:
- `Filtered_AlnLength_Distr.svg`
- `Matches.svg`
- `Scatterplot_Aln.png`
- `Histogram_Aln_Start_End.svg`
- `Count_comparison_matches.svg`
- `Startpoints2000bp.svg`

---

## 1. Alignment to Transgene References

FASTQ files are aligned to the respective transgene reference sequence
using minimap2.


```{bash alignment, eval=FALSE}
# Input:  FASTQ files per sample, reference FASTA per transgene variant
# Output: Sorted, indexed BAM files

SAMPLES=("lambda" "wt-fragments" "M1" "M2" "M3" "M4")
FASTQ_DIR="/path"
REF_DIR="/path"
ALN_DIR="/path"

mkdir -p "${ALN_DIR}"

for SAMPLE in "${SAMPLES[@]}"; do
    minimap2 -ax map-ont \
        "${REF_DIR}/transgene_cassette_${SAMPLE}.fasta" \
        "${FASTQ_DIR}/${SAMPLE}.fastq" \
        | samtools view -bS \
        | samtools sort -o "${ALN_DIR}/sorted_alignment_${SAMPLE}.bam"

    samtools index "${ALN_DIR}/sorted_alignment_${SAMPLE}.bam"
done
```

---

## 2. Filter BAM Files

Secondary (flag 256), supplementary (flag 2048) and unmapped (flag 4)
alignments are removed. Combined flag: -F 2308.


```{bash filter, eval=FALSE}
# Input:  Sorted BAM files
# Output: Filtered BAM files (primary alignments only)

ALN_DIR="/path"
FILTERED_DIR="/path"

mkdir -p "${FILTERED_DIR}"

for SAMPLE in "${SAMPLES[@]}"; do
    samtools view -F 2308 -b -h \
        "${ALN_DIR}/sorted_alignment_${SAMPLE}.bam" \
        > "${FILTERED_DIR}/filtered_supp_unm_sec_${SAMPLE}.bam"

    samtools index "${FILTERED_DIR}/filtered_supp_unm_sec_${SAMPLE}.bam"
done
```

---

## 3. Load Filtered BAM Files

```{r load_bam}
filtered_files <- list.files(
  bam_dir,
  pattern    = "\\.bam$",
  full.names = TRUE,
  recursive  = TRUE
)

sample_names_f <- basename(filtered_files) %>%
  gsub("filtered_supp_unm_sec_", "", .) %>%
  gsub("\\.bam$", "", .)

# Load all BAM files into AlignedBamTable objects
filtered_objects <- lapply(filtered_files, loadAlignedBamTable)
names(filtered_objects) <- sample_names_f

# Combine into single data frame
aln_df <- lapply(filtered_objects, function(obj) obj@table) %>%
  bind_rows() %>%
  mutate(
    barcode       = gsub("filtered_supp_unm_sec_", "", barcode) %>%
                    gsub("sorted_alignment_", "", .),
    barcode       = factor(barcode, levels = sample_levels),
    strand_label  = ifelse(strand == "1", "+", "-"),
    aln_start_new = ifelse(strand == "1", aln_start, aln_end),
    aln_end_new   = ifelse(strand == "1", aln_end,   aln_start)
  )
```

---

## 4. Alignment Length Distribution

```{r aln_length}
aln_length_df <- aln_df %>% select(Sample = barcode, aln_length)

gghistogram(
  data       = aln_length_df,
  x          = "aln_length",
  fill       = "Sample",
  color      = "black",
  bins       = 20,
  add        = "mean",
  add.params = list(color = "black", linetype = "dashed", linewidth = 0.25),
  rug        = FALSE
) +
  scale_fill_manual(values = sample_colors) +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +
  coord_cartesian(xlim = c(0, 8000)) +
  facet_wrap(~ Sample, ncol = 3) +
  labs(title = "Alignment Length Distribution per Sample",
       x = "Alignment Length (bp)", y = "Count") +
  theme_publication(base_size = 10) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(file.path(output_dir, "Filtered_AlnLength_Distr.svg"),
       width = 180, height = 120, units = "mm")
```

---

## 5. Match Count Distribution

```{r match_count}
match_df <- lapply(seq_along(filtered_files), function(i) {
  reads <- readGAlignments(filtered_files[i])
  cig   <- cigar(reads)
  match_counts <- sapply(cig, function(c) {
    lengths <- as.numeric(unlist(regmatches(c, gregexpr("[0-9]+", c))))
    ops     <- unlist(regmatches(c, gregexpr("[A-Z]", c)))
    sum(lengths[ops == "M"])
  })
  data.frame(Sample = sample_names_f[i], matches = match_counts)
}) %>% bind_rows() %>%
  mutate(Sample = factor(Sample, levels = sample_levels))

gghistogram(
  data  = match_df,
  x     = "matches",
  fill  = "Sample",
  color = "transparent",
  bins  = 50,
  alpha = 1
) +
  scale_fill_manual(values = sample_colors) +
  scale_y_continuous(
    limits = c(0, 125000),
    breaks = seq(0, 100000, by = 50000),
    labels = function(x) format(x, scientific = FALSE)
  ) +
  coord_cartesian(xlim = c(0, 8000)) +
  facet_wrap(~ Sample, ncol = 3) +
  labs(title = "Match Distribution per Sample",
       x = "Number of Matches (M)", y = "Count") +
  theme_publication(base_size = 10) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(file.path(output_dir, "Matches.svg"),
       width = 180, height = 120, units = "mm")
```

---

## 6. Alignment Start vs. End Scatterplot

```{r scatterplot}
ggscatter(
  data    = aln_df,
  x       = "aln_start_new",
  y       = "aln_end_new",
  color   = "strand_label",
  palette = c("+" = "#F44336", "-" = "#2196F3"),
  size    = 0.1,
  alpha   = 0.3
) +
  scale_x_continuous(
    breaks = seq(0, max(aln_df$aln_start_new, na.rm = TRUE), by = 2000)
  ) +
  facet_wrap(~ barcode, ncol = 3) +
  labs(x = "Alignment Start", y = "Alignment End", color = "Strand") +
  theme_publication(base_size = 10) +
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(file.path(output_dir, "Scatterplot_Aln.png"),
       width = 180, height = 120, units = "mm", dpi = 600)
```

---

## 7. Histogram of Alignment Start and End Positions

```{r aln_histogram}
aln_df_div <- aln_df %>%
  mutate(
    aln_start_new = ifelse(strand == "-1", aln_end,   aln_start),
    aln_end_new   = ifelse(strand == "-1", aln_start, aln_end)
  )

df_long <- aln_df_div %>%
  pivot_longer(
    cols      = c(aln_start_new, aln_end_new),
    names_to  = "variable",
    values_to = "coordinate"
  )

ggplot(df_long, aes(x = coordinate, fill = variable)) +
  geom_histogram(
    data     = df_long %>% filter(strand == "1"),
    alpha    = 1, bins = 50, position = "identity"
  ) +
  geom_histogram(
    data     = df_long %>% filter(strand == "-1"),
    aes(y    = -after_stat(count)),
    alpha    = 1, bins = 50, position = "identity"
  ) +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +
  scale_fill_manual(
    values = c("aln_start_new" = "black", "aln_end_new" = "#CD6600"),
    labels = c("aln_start_new" = "Aln Start", "aln_end_new" = "Aln End")
  ) +
  geom_hline(yintercept = 0) +
  facet_wrap(~ barcode, ncol = 3) +
  labs(x = "Position [bp]", y = "Count", fill = "Variable") +
  theme_publication(base_size = 10) +
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(file.path(output_dir, "Histogram_Aln_Start_End.svg"),
       width = 180, height = 120, units = "mm")
```

---

## 8. Genome Integrity (Complete vs. Truncated)

```{r genome_integrity}
genome_length <- 4700

match_df <- match_df %>%
  mutate(
    genome_type = case_when(
      matches >= genome_length * 0.95 &
        matches <= genome_length * 1.05 ~ "Complete (±5%)",
      TRUE ~ "Truncated"
    )
  )

genome_stats <- match_df %>%
  group_by(Sample, genome_type) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(Sample) %>%
  mutate(
    percent     = round(count / sum(count) * 100, 1),
    genome_type = factor(genome_type, levels = c("Complete (±5%)", "Truncated"))
  )

ggbarplot(
  data          = genome_stats,
  x             = "Sample",
  y             = "percent",
  fill          = "genome_type",
  color         = "transparent",
  palette       = c("Complete (±5%)" = "#8FBC8F", "Truncated" = "#FF7256"),
  label         = TRUE,
  lab.nb.digits = 1,
  lab.size      = 3,
  lab.vjust     = 0.5
) +
  scale_y_continuous(
    labels = function(x) paste0(x, "%"),
    expand = expansion(mult = c(0.05, 0.1))
  ) +
  labs(x = "Sample", y = "Percentage of Reads (%)", fill = "Genome Type") +
  theme_publication(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right")

ggsave(file.path(output_dir, "Count_comparison_matches.svg"),
       width = 180, height = 120, units = "mm")
```

---

## 9. Detail View: Alignment Start Positions at ~2000 bp

```{r starts_2000bp}
reads_2000 <- aln_df %>%
  filter(aln_start >= 1800 & aln_start <= 2200)

cat("Number of reads with start position at ~2000 bp:", nrow(reads_2000), "\n")
summary(reads_2000$read_len)
summary(reads_2000$aln_length)
table(reads_2000$strand_label)

gghistogram(
  data  = reads_2000,
  x     = "aln_start",
  fill  = "strand_label",
  color = "transparent",
  bins  = 100
) +
  scale_fill_manual(values = c("+" = "#F44336", "-" = "#2196F3")) +
  facet_wrap(~ barcode, ncol = 3) +
  labs(title = "Start Positions Distribution at ~2000 bp",
       x = "Alignment Start", y = "Count", fill = "Strand") +
  theme_publication(base_size = 10)

ggsave(file.path(output_dir, "Startpoints2000bp.svg"),
       width = 180, height = 120, units = "mm")
```

---


