---
title: "04 Coverage and InDel Analysis"
author: "Joyce Suppacher"
date: "`2026-06-14"
---

# Coverage and Indel Analysis

This script visualizes coverage and InDel profiles across the vector genome
for all six rAAV2 cassettes, as well as a separate detailed analysis of the
wt-fragments cassette.

Approach:
- Coverage is calculated from filtered BAM files (primary alignments only)
- Insertions, deletions, Soft-clip 5' and Soft-clip 3' are extracted from CIGAR strings
- InDel counts are normalized to total read count per sample to allow
  comparison across samples with different sequencing depths
- PolyA homopolymer regions (≥5 consecutive adenines) are excluded from
  InDel analysis to avoid sequencing artifacts

Run `00_setup.Rmd` first.

```{r packages, message=FALSE, warning=FALSE}
library(GenomicAlignments) # readGAlignments(), coverage(), cigar(), start(), strand()
library(GenomicRanges)     # GRanges(), IRanges()
library(Biostrings)        # readDNAStringSet()
library(dplyr)             # mutate(), filter(), group_by(), summarise(), bind_rows()
library(purrr)             # map2()
library(ggplot2)           # ggplot(), geom_line(), scale_color_manual(), ggsave()
```

Outputs:
- `Coverage_plot.svg`: Coverage + InDel plot for all six variants
- `Coverage_wt-fragments.svg`: Detailed coverage + InDel plot for wt-fragments

---

## 1. Load Filtered BAM Files and Reference Sequences

```{r load_data}
# Load filtered BAM files
filtered_files <- list.files(
  bam_dir,
  pattern    = "\\.bam$",
  full.names = TRUE,
  recursive  = TRUE
)

names(filtered_files) <- basename(filtered_files) %>%
  gsub("filtered_supp_unm_sec_", "", .) %>%
  gsub("\\.bam$", "", .)

# Enforce sample order
filtered_files <- filtered_files[sample_levels]

# Load reference FASTA files for PolyA detection
fasta_files_ref <- list.files(
  ref_dir,
  pattern    = "\\.fasta$|\\.fa$",
  full.names = TRUE
)

fasta_names_ref <- basename(fasta_files_ref) %>%
  gsub("\\.fasta$|\\.fa$", "", .) %>%
  gsub("transgene_cassette_", "", .)
```

---

## 2. Detect PolyA Homopolymer Regions

PolyA regions (≥5 consecutive adenines) are identified in each reference
sequence and excluded from InDel analysis to avoid sequencing artifacts.

```{r polya}
polyA_list <- map2(fasta_files_ref, fasta_names_ref, function(fasta_file, sample_name) {
  ref      <- readDNAStringSet(fasta_file)
  polyA_df <- lapply(names(ref), function(seq_name) {
    seq     <- ref[[seq_name]]
    matches <- gregexpr("A{5,}", as.character(seq))[[1]]
    if (matches[1] == -1) return(NULL)
    data.frame(
      seqname = seq_name,
      start   = as.integer(matches),
      end     = as.integer(matches) + attr(matches, "match.length") - 1
    )
  }) %>% bind_rows()

  GRanges(
    seqnames = polyA_df$seqname,
    ranges   = IRanges(start = polyA_df$start, end = polyA_df$end)
  )
})
names(polyA_list) <- fasta_names_ref


global_xmax <- max(unlist(lapply(names(filtered_files), function(s) {
  bam <- readGAlignments(filtered_files[[s]])
  max(as.integer(end(bam)), na.rm = TRUE)
})))
```

---

## 3. Coverage and Indel Extraction Function

Extracts per-position coverage, deletions, insertions, Soft-clip 5' and 3'
from a GAlignments object. InDels are normalized to total read count.

```{r extract_function}
extract_cov_indels <- function(bam_file, sample_name, polyA_gr = NULL) {

  bam         <- bam_file
  vektor_name <- names(seqlengths(bam))[1]
  num_seqs    <- length(bam)   # total read count for normalization

 # Coverage
  cov_vec <- as.numeric(coverage(bam)[[vektor_name]])
  coverage_df <- data.frame(
    sample = sample_name,
    pos    = seq_along(cov_vec),
    count  = cov_vec,
    type   = "coverage",
    strand = "+"
  )

  # PolyA positions
  polyA_positions <- integer()
  if (!is.null(polyA_gr)) {
    polyA_positions <- unlist(
      lapply(seq_along(polyA_gr), function(i) start(polyA_gr[i]):end(polyA_gr[i]))
    )
  }

  # Extract InDels and Soft-clips from CIGAR strings
  cigars  <- cigar(bam)
  starts  <- start(bam)
  strands <- as.character(strand(bam))
  res     <- vector("list", length(cigars))

  for (i in seq_along(cigars)) {
    cig        <- cigars[i]
    ref_pos    <- starts[i]
    strand_i   <- strands[i]
    lengths    <- as.numeric(unlist(regmatches(cig, gregexpr("[0-9]+", cig))))
    ops        <- unlist(regmatches(cig, gregexpr("[A-Z]", cig)))
    pos_vec    <- c()
    type_vec   <- c()
    strand_vec <- c()

    for (j in seq_along(ops)) {
      op  <- ops[j]
      len <- lengths[j]

      if (op == "D") {
        # Deletion: spans multiple reference positions
        pos_vec    <- c(pos_vec, ref_pos:(ref_pos + len - 1))
        type_vec   <- c(type_vec, rep("del", len))
        strand_vec <- c(strand_vec, rep(strand_i, len))
        ref_pos    <- ref_pos + len

      } else if (op == "I") {
        # Insertion: single reference position
        pos_vec    <- c(pos_vec, ref_pos)
        type_vec   <- c(type_vec, "ins")
        strand_vec <- c(strand_vec, strand_i)

      } else if (op == "S") {
        # SoftClip: first S = 5', last S = 3'
        clip_type  <- if (j == 1) "SoftClip_5" else "SoftClip_3"
        pos_vec    <- c(pos_vec, ref_pos)
        type_vec   <- c(type_vec, clip_type)
        strand_vec <- c(strand_vec, strand_i)

      } else if (op %in% c("M", "=", "X")) {
        ref_pos <- ref_pos + len
      }
    }

    if (length(pos_vec) > 0) {
      res[[i]] <- data.frame(pos = pos_vec, type = type_vec, strand = strand_vec)
    }
  }

  indel_df <- bind_rows(res)

  # Exclude PolyA positions
  if (length(polyA_positions) > 0) {
    indel_df <- indel_df %>% filter(!pos %in% polyA_positions)
  }

 
  indel_counts <- indel_df %>%
    group_by(pos, type, strand) %>%
    summarise(count = n(), .groups = "drop") %>%
    mutate(
      sample = sample_name,
      count  = ifelse(strand == "-", -(count / num_seqs), count / num_seqs)
    )

  bind_rows(coverage_df, indel_counts)
}
```

---

## 4. Run Extraction for All Samples

```{r extract_all}
cov_all <- lapply(names(filtered_files), function(sample_name) {
  bam      <- readGAlignments(filtered_files[[sample_name]])
  polyA_gr <- if (sample_name %in% names(polyA_list)) polyA_list[[sample_name]] else NULL
  extract_cov_indels(bam, sample_name, polyA_gr)
}) %>%
  bind_rows() %>%
  mutate(sample = factor(sample, levels = sample_levels))
```

---

## 5. Coverage + Indel Plot – All Samples

Secondary y-axis shows normalized InDel/Soft-clip signal.


```{r coverage_plot_all}
max_coverage   <- max(cov_all %>% filter(type == "coverage") %>% pull(count), na.rm = TRUE)
primary_breaks <- seq(0, max_coverage, by = 50000)
scale_factor   <- 50000 / 0.1   # aligns coverage and indel axes

p_all <- ggplot(cov_all, aes(x = pos, color = type)) +
  geom_line(
    data      = cov_all %>% filter(type == "coverage"),
    aes(y     = count),
    linewidth = 1
  ) +
  geom_line(
    data      = cov_all %>% filter(type != "coverage"),
    aes(y     = count * scale_factor, linetype = strand),
    linewidth = 0.2,
    alpha     = 0.8
  ) +
  geom_hline(yintercept = 0, linewidth = 0.3, color = "grey50") +
  scale_color_manual(
    name   = "Signal",
    values = c(
      "coverage"   = "black",
      "del"        = "red",
      "ins"        = "blue",
      "SoftClip_5" = "green",
      "SoftClip_3" = "darkgreen"
    )
  ) +
  scale_linetype_manual(
    name  = "Strand",
    values = c("+" = "solid", "-" = "solid"),
    guide = "none"
  ) +
  scale_y_continuous(
    name     = "Coverage",
    breaks   = primary_breaks,
    limits   = c(-max_coverage * 0.1, max_coverage),
    sec.axis = sec_axis(
      ~ . / scale_factor,
      name   = "Relative InDels",
      breaks = primary_breaks / scale_factor
    )
  ) +
  scale_x_continuous(
    limits = c(0, global_xmax),
    expand = expansion(mult = c(0, 0.02))
  ) +
  facet_wrap(~ sample, ncol = 3) +
  labs(x = "Position (bp)") +
  theme_publication(base_size = 10) +
  theme(legend.position = "right")

p_all

ggsave(file.path(output_dir, "Coverage_plot.svg"), plot = p_all,
       width = 180, height = 120, units = "mm")
```

---

## 6. Coverage + Indel Plot (wt-fragments)

Separate detailed analysis of the wt-fragments sample.

```{r coverage_wtf}

cov_wtf <- cov_all %>% filter(sample == "wt-fragments")

max_coverage_wtf   <- max(cov_wtf %>% filter(type == "coverage") %>% pull(count), na.rm = TRUE)
primary_breaks_wtf <- seq(0, max_coverage_wtf, by = 50000)
scale_factor_wtf   <- 50000 / 0.1

p_wtf <- ggplot(cov_wtf, aes(x = pos, color = type)) +
  geom_line(
    data      = cov_wtf %>% filter(type == "coverage"),
    aes(y     = count),
    linewidth = 1
  ) +
  geom_line(
    data      = cov_wtf %>% filter(type != "coverage"),
    aes(y     = count * scale_factor_wtf, linetype = strand),
    linewidth = 0.2,
    alpha     = 0.8
  ) +
  geom_hline(yintercept = 0, linewidth = 0.3, color = "grey50") +
  scale_color_manual(
    name   = "Signal",
    values = c(
      "coverage"   = "black",
      "del"        = "red",
      "ins"        = "blue",
      "SoftClip_5" = "green",
      "SoftClip_3" = "darkgreen"
    )
  ) +
  scale_linetype_manual(
    name  = "Strand",
    values = c("+" = "solid", "-" = "solid"),
    guide = "none"
  ) +
  scale_y_continuous(
    name     = "Coverage",
    breaks   = primary_breaks_wtf,
    limits   = c(-max_coverage_wtf * 0.1, max_coverage_wtf),
    sec.axis = sec_axis(
      ~ . / scale_factor_wtf,
      name   = "Relative InDels",
      breaks = primary_breaks_wtf / scale_factor_wtf
    )
  ) +
  scale_x_continuous(
    limits = c(0, global_xmax),
    expand = expansion(mult = c(0, 0.02))
  ) +
  labs(
    title = "Coverage wt-fragments",
    x     = "Position (bp)"
  ) +
  theme_publication(base_size = 10) +
  theme(legend.position = "right")

p_wtf

ggsave(file.path(output_dir, "Coverage_wt-fragments.svg"), plot = p_wtf,
       width = 180, height = 120, units = "mm")
```

---
