---
title: "05 Contamination Analysis"
author: "Joyce Suppacher"
date: "2026-06-14"
---

# Contamination Analysis

This script covers:
1. Alignment to cap/rep reference sequences (PAF files)
2. Extraction of contaminating read names (R)
3. Subsetting FASTQ to contaminating reads (seqtk)
4. Re-alignment to virus reference (BAM)
5. Metagenome construction and alignment
6. Contamination plot

Run `00_setup.Rmd` first.

```{r packages, message=FALSE, warning=FALSE}
library(Rsamtools)         # countBam()
library(dplyr)             # mutate(), filter(), distinct(), bind_rows(), left_join()
library(readr)             # read_tsv()
library(ggpubr)            # ggbarplot()
library(ggplot2)           # scale_y_continuous(), ggsave()
library(gt)                # gt(), tab_header(), cols_label()
```

Output:
- `contamination_percent.svg`

---

## 1. Align to cap/rep Reference Sequences

FASTQ files are aligned to individual cap/rep reference sequences
to identify reads containing contaminating sequences.


```{bash paf_alignment, eval=FALSE}
# Input:  FASTQ files per sample, FASTA per contaminant (VP1, VP3, Rep78)
# Output: PAF files per sample/contaminant combination

SAMPLES=("lambda" "wt-fragments" "M1" "M2" "M3" "M4")
CONTAMINANTS=("VP1" "VP3" "rep78")
FASTQ_DIR="/path"
REF_DIR="/path"
CONT_DIR="/path"

mkdir -p "${CONT_DIR}"

for SAMPLE in "${SAMPLES[@]}"; do
    for CONT in "${CONTAMINANTS[@]}"; do
        minimap2 -x map-ont \
            "${REF_DIR}/AAV_${CONT}.fasta" \
            "${FASTQ_DIR}/${SAMPLE}.fastq" \
            > "${CONT_DIR}/${CONT}_${SAMPLE}.paf"
    done
done
```

---

## 2. Extract Contaminating Read Names

Read names of full-length alignments are extracted from PAF files in R
and saved as TXT files for use with seqtk.

```{r extract_readnames}
# Example for VP1 (repeat analogously for VP3 and Rep78)
paf_dir  <- file.path(cont_dir, "contamination_reads")
cont_dir_reads <- file.path(cont_dir, "contamination_reads")

extract_readnames <- function(paf_file, min_length = 1000) {
  paf <- read_tsv(paf_file,
    col_names = c("qname", "qlen", "qstart", "qend", "strand",
                  "tname", "tlen", "tstart", "tend", "matches",
                  "aln_len", "mapq"),
    col_types = cols(.default = "c")
  ) %>%
    mutate(aln_len = as.numeric(aln_len)) %>%
    filter(aln_len >= min_length)
  unique(paf$qname)
}

for (sample in sample_levels) {
  for (cont in c("VP1", "VP3", "rep78")) {
    paf_file <- file.path(cont_dir_reads, paste0(cont, "_", sample, ".paf"))
    if (file.exists(paf_file)) {
      readnames <- extract_readnames(paf_file)
      out_file  <- file.path(cont_dir_reads,
                             paste0("readnames_", cont, "_", sample, ".txt"))
      writeLines(readnames, out_file)
      cat(sample, cont, "–", length(readnames), "reads\n")
    }
  }
}
```

---

## 3. Subset FASTQ to Contaminating Reads

Contaminating reads are extracted from FASTQ files using seqtk.
Note: eval=FALSE – run on HPC.

```{bash seqtk_subset, eval=FALSE}
# Input:  FASTQ files, TXT files with read names
# Output: FASTQ files containing only contaminating reads

SAMPLES=("lambda" "wt-fragments" "M1" "M2" "M3" "M4")
CONTAMINANTS=("VP1" "VP3" "rep78")
FASTQ_DIR="/your/path/data/02_fastq/A_tailing"
CONT_DIR="/your/path/data/08_contaminations/contamination_reads"

for SAMPLE in "${SAMPLES[@]}"; do
    for CONT in "${CONTAMINANTS[@]}"; do
        READNAMES="${CONT_DIR}/readnames_${CONT}_${SAMPLE}.txt"
        if [ -f "${READNAMES}" ]; then
            seqtk subseq \
                "${FASTQ_DIR}/${SAMPLE}.fastq" \
                "${READNAMES}" \
                > "${CONT_DIR}/${CONT}_${SAMPLE}.fastq"
        fi
    done
done
```

---

## 4. Re-align Contaminating Reads to Virus Reference

```{bash realignment, eval=FALSE}
# Input:  FASTQ files with contaminating reads
# Output: Sorted, indexed BAM files

SAMPLES=("lambda" "wt-fragments" "M1" "M2" "M3" "M4")
CONTAMINANTS=("VP1" "VP3" "rep78")
CONT_DIR="/path"
VIRUS_REF="/path"

mkdir -p "${CONT_DIR}/contamination_alignments"

for SAMPLE in "${SAMPLES[@]}"; do
    for CONT in "${CONTAMINANTS[@]}"; do
        FASTQ="${CONT_DIR}/contamination_reads/${CONT}_${SAMPLE}.fastq"
        BAM_OUT="${CONT_DIR}/contamination_alignments/${CONT}_${SAMPLE}.sorted.bam"
        if [ -f "${FASTQ}" ]; then
            minimap2 -ax map-ont "${VIRUS_REF}" "${FASTQ}" \
                | samtools view -bS \
                | samtools sort -o "${BAM_OUT}"
            samtools index "${BAM_OUT}"
        fi
    done
done
```

---

## 5. Build Metagenome and Align

The metagenome combines all contamination reference sequences:
generic backbone, AdH sequences, cap/rep-specific sequences (WHc2), and hg38.


```{bash metagenome, eval=FALSE}
# Build metagenome
CONT_REF_DIR="/path"
METAGENOME="/path"

mkdir -p "$(dirname ${METAGENOME})"
cat "${CONT_REF_DIR}"/*.fasta > "${METAGENOME}"

# Align each sample to metagenome
SAMPLES=("lambda" "wt-fragments" "M1" "M2" "M3" "M4")
FASTQ_DIR="/path"
GB_DIR="/path"

mkdir -p "${GB_DIR}"

for SAMPLE in "${SAMPLES[@]}"; do
    minimap2 -ax map-ont "${METAGENOME}" "${FASTQ_DIR}/${SAMPLE}.fastq" \
        | samtools view -bS \
        | samtools sort -o "${GB_DIR}/${SAMPLE}_gb.sorted.bam"
    samtools index "${GB_DIR}/${SAMPLE}_gb.sorted.bam"
done
```

---

## 6. Load and Process Contamination Data

```{r load_cont}
load_cont <- function(sample_name, file_suffix) {
  file <- file.path(cont_dir, "generic_backbone",
                    paste0(file_suffix, ".sorted.bam"))
  obj  <- loadAlignedBamTable(file)
  obj@table %>%
    filter(mapq >= 50) %>%
    mutate(
      seqnames_grouped = case_when(
        grepl("backbone|AdH|WHc2", seqnames) ~ seqnames,
        TRUE ~ "Human Genome"
      )
    ) %>%
    distinct(qname, seqnames_grouped) %>%
    dplyr::count(seqnames_grouped, sort = TRUE)
}

cont_gb_lambda <- load_cont("lambda",       "lambda_gb")
cont_gb_wtf    <- load_cont("wt-fragments", "wt-fragment_gb")
cont_gb_M1     <- load_cont("M1",           "M1_gb")
cont_gb_M2     <- load_cont("M2",           "M2_gb")
cont_gb_M3     <- load_cont("M3",           "M3_gb")
cont_gb_M4     <- load_cont("M4",           "M4_gb")
```

---

## 7. Calculate Total Reads and Percentages

```{r total_reads}
fastq_files <- list.files(
  fastq_dir,
  pattern    = "\\.fastq$|\\.fastq\\.gz$",
  full.names = TRUE
)

total_reads <- data.frame(
  Sample = basename(fastq_files) %>%
             gsub("\\.fastq\\.gz$|\\.fastq$", "", .),
  total  = sapply(fastq_files, function(f) {
    length(readDNAStringSet(f, format = "fastq"))
  })
)

cont_all <- bind_rows(
  cont_gb_lambda %>% mutate(Sample = "lambda"),
  cont_gb_wtf    %>% mutate(Sample = "wt-fragments"),
  cont_gb_M1     %>% mutate(Sample = "M1"),
  cont_gb_M2     %>% mutate(Sample = "M2"),
  cont_gb_M3     %>% mutate(Sample = "M3"),
  cont_gb_M4     %>% mutate(Sample = "M4")
) %>%
  left_join(total_reads, by = "Sample") %>%
  mutate(
    percent          = round(n / total * 100, 6),
    Sample           = factor(Sample, levels = sample_levels),
    seqnames_grouped = factor(seqnames_grouped,
                              levels = c("backbone", "AdH", "cap/rep", "Human Genome"))
  )
```

---

## 8. Contamination Plot

```{r cont_plot}
ggbarplot(
  data    = cont_all,
  x       = "Sample",
  y       = "percent",
  fill    = "seqnames_grouped",
  color   = "transparent",
  palette = c(
    "backbone"     = "#4CAF50",
    "AdH"          = "#F44336",
    "cap/rep"      = "#FF9800",
    "Human Genome" = "#2196F3"
  )
) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    title = "Contamination per Sample",
    x     = "Sample",
    y     = "% of Total Reads",
    fill  = "Category"
  ) +
  theme_publication(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right")

ggsave(file.path(output_dir, "contamination_percent.svg"),
       width = 90, height = 60, units = "mm")
```

---

