---
title: "06 ITR Distance Analysis"
author: "Joyce Suppacher"
date: "2026-06-14"
---

# ITR Distance Analysis

This script covers:
1. Conversion of contaminating FASTQ to FASTA
2. BLAST of ITR sequences against contaminating reads
3. Classification of reads as "correctly spaced" (ITR-to-ITR distance >= 2000 bp)

Threshold of 2000 bp reflects minimum expected length of ITR-flanked read
containing the shortest contaminant (Rep78, ~1900 bp + 2x ITR ~130 bp).

Run `00_setup.Rmd` and `05_contamination.Rmd` first.

```{r packages, message=FALSE, warning=FALSE}
library(readr)    # read_tsv(): load BLAST output files
library(dplyr)    # mutate(), group_by(), summarise(), bind_rows(), left_join()
library(purrr)    # map_dfr(): load multiple BLAST files
library(gt)       # gt(), tab_header(), cols_label(), tab_style()
```

Output:
- ITR distance summary table (gt)

---

## 1. Convert Contaminating FASTQ to FASTA

Required for BLAST input format.
Note: eval=FALSE – run on HPC.

```{bash fastq_to_fasta, eval=FALSE}
# Input:  FASTQ files with contaminating reads (from 04_contamination.Rmd Step 3)
# Output: FASTA files per sample/contaminant

SAMPLES=("lambda" "wt-fragments" "M1" "M2" "M3" "M4")
CONTAMINANTS=("VP1" "VP3" "rep78")
CONT_DIR="/path"
BLAST_DIR="/path"

mkdir -p "${BLAST_DIR}/VP1" "${BLAST_DIR}/VP3" "${BLAST_DIR}/rep78"

for SAMPLE in "${SAMPLES[@]}"; do
    for CONT in "${CONTAMINANTS[@]}"; do
        FASTQ="${CONT_DIR}/${CONT}_${SAMPLE}.fastq"
        FASTA="${BLAST_DIR}/${CONT}/${CONT}_${SAMPLE}.fasta"
        if [ -f "${FASTQ}" ]; then
            seqtk seq -A "${FASTQ}" > "${FASTA}"
        fi
    done
done
```

---

## 2. BLAST ITR Sequences Against Contaminating Reads

Both AAV2 ITRs (ITR1 and ITR2) are used as queries.


```{bash blast, eval=FALSE}
# Input:  ITR FASTA (query), contaminating read FASTA (subject)
# Output: TSV files with BLAST hits

ITR_FASTA="/path"
BLAST_DIR="/path"

for SAMPLE in "${SAMPLES[@]}"; do
    for CONT in "${CONTAMINANTS[@]}"; do
        FASTA="${BLAST_DIR}/${CONT}/${CONT}_${SAMPLE}.fasta"
        TSV_OUT="${BLAST_DIR}/${CONT}/ITR_${CONT}_${SAMPLE}.tsv"
        if [ -f "${FASTA}" ]; then
            blastn \
                -query   "${ITR_FASTA}" \
                -subject "${FASTA}" \
                -out     "${TSV_OUT}" \
                -outfmt  "6 qseqid sseqid pident length mismatch gapopen qstart qend sstart send evalue bitscore"
        fi
    done
done
```

---

## 3. Load BLAST Results

```{r load_blast}
samples_vp1 <- c("M1", "M2", "M4")

all_vp1 <- map_dfr(samples_vp1, ~ {
  read_tsv(
    file.path(blast_dir, "VP1", paste0("ITR_VP1_", .x, ".tsv")),
    col_names = c("qseqid", "sseqid", "pident", "length", "mismatch",
                  "gapopen", "qstart", "qend", "sstart", "send",
                  "evalue", "bitscore")
  ) %>%
    filter(length >= 130) %>%
    mutate(
      contaminants = "VP1",
      Sample       = .x,
      itr_type     = case_when(
        qseqid == "AAV2_ITR1" ~ "ITR1",
        qseqid == "AAV2_ITR2" ~ "ITR2",
        TRUE ~ qseqid
      ),
      group = paste0(Sample, "_", itr_type)
    )
})

# Load VP3 and Rep78 analogously
# all_vp3      <- map_dfr(samples_vp3, ~ { ... })
# itr_data_rep <- map_dfr(samples_rep, ~ { ... })
```

---

## 4. Classify Reads as Correctly Spaced

Reads with ITR-to-ITR distance >= 2000 bp are classified as
"correctly spaced".

```{r classify}
all_data <- bind_rows(
  itr_data_rep,
  all_vp1,
  all_vp3
) %>%
  mutate(pos = pmin(sstart, send))

result <- all_data %>%
  group_by(contaminants, Sample, sseqid) %>%
  summarise(
    n_hits  = n(),
    max_pos = max(pos),
    min_pos = min(pos),
    distant = if (n_hits < 2) {
      FALSE
    } else {
      (max_pos - min_pos) >= 2000
    },
    .groups = "drop"
  )
```

---

## 5. Summarise and Display Results

```{r summarise}
summary_itr <- result %>%
  group_by(contaminants, Sample) %>%
  summarise(
    total_reads   = n(),
    distant_reads = sum(distant),
    percent       = 100 * distant_reads / total_reads,
    .groups       = "drop"
  )

final <- summary_itr %>%
  left_join(read_stats, by = "Sample") %>%
  mutate(percent_of_library = distant_reads / Raw * 100) %>%
  select(-Mapped, -Unmapped, -Filtered, -Raw,
         -Mapped_pct, -Unmapped_pct, -Filtered_pct, -percent)

final %>%
  gt() %>%
  tab_header(title = "ITR Distance") %>%
  cols_label(
    contaminants       = "Contaminant",
    Sample             = "Sample",
    total_reads        = "ITR-containing reads",
    distant_reads      = "Correctly spaced ITR reads",
    percent_of_library = "% of Library"
  ) %>%
  cols_align(align = "center", columns = everything()) %>%
  tab_style(style = cell_text(color = "black"), locations = cells_body()) %>%
  tab_style(style = cell_text(color = "black"), locations = cells_column_labels()) %>%
  tab_style(style = cell_text(color = "black"), locations = cells_title())
```

---
